Inferring community characteristics in labelled networks

### 2021-05-23 nips/conclusion.tex
Finish draft

The FFBM was developed to address the shortcomings of other graphical models when testing how vertex features affect community structure. The idea is to divide the graph into its most natural partition and test whether the vertex features can accurately explain this partition. We therefore hope to avoid a common pitfall in data science: correlation does not imply causation. It is very easy to find vertex features that are in some way correlated with the graphical structure. Nonetheless, only when we find the feature that best describes the most pronounced partition do we have a stronger case for causation.

With the newly-defined FFBM, we go on to present an efficient inference algorithm to sample the parameters $\theta$ of the feature-to-block generator. This is introduced as two concurrent Markov chains to sample the block memberships $b$ and block generator parameters $\theta$. The key trick to make the $b$-chain efficient is the use of the microcanonical rather than the traditional SBM (as proposed by \citet{Peixoto-Bayesian-Microcanonical}) which allows us to bypass an expensive summation over the latent memberships $b$. Additionally, we use the empirical mean of the $b$-samples as the input to our $\theta$-chain, as this reduces the variance in our evaluation of the target distribution and thus shortens the burn-in and auto-correlation times of our $\theta$-chain.

The overall method is shown to be effective at extracting and describing the most natural communities in a labelled network. Nevertheless, the approach can only currently explain the structure at the macro-scale. We cannot explain structure within each community. Future work will benefit from extending the FFBM to be hierarchical in nature. That way, the structure of the network can be explained at all length-scales of interest. This work may be built upon to develop better methods of analysing structure in networks and drawing evidence-backed conclusions. So long as data collection techniques remain ethical and care is taken to respect personal privacy, such empowered decision-making can only help humankind.

>>>>

The FFBM was developed to address the shortcomings of other graphical models when testing how vertex features affect community structure. The idea is to divide the graph into its most natural partition and test whether the vertex features can accurately explain this partition. It is very easy to find vertex features that are in some way correlated with the graphical structure. Nonetheless, only when we find the feature that best describes the most pronounced partition do we have a stronger case for causation.

With the newly-defined FFBM, we go on to present an efficient inference algorithm to sample the parameters $\theta$ of the feature-to-block generator. This is introduced as two concurrent Markov chains to sample the block memberships $b$ and block generator parameters $\theta$. Nevertheless, we can serialise the chains and use the empirical mean of the $b$-samples as the input to our $\theta$-chain. This reduces the variance in our evaluation of the target distribution and thus shortens burn-in.

The overall method is shown to be effective at extracting and describing the most natural communities in a labelled network. Nevertheless, the approach can only currently explain the structure at the macro-scale. We cannot explain structure within each block. Future work will benefit from extending the FFBM to be hierarchical in nature. That way, the structure of the network can be explained at all length-scales of interest. So long as data collection techniques remain ethical and care is taken to respect personal privacy, such empowered decision-making can only help humankind.

### 2021-05-23 nips/experiments.tex
Finish draft

We apply the developed methods to a variety of datasets. Datasets are chosen to span a range of node counts $N$, edge counts $E$ and feature space dimension $D$. We consider the following datasets:\footnote{For information as to how these data were collected and how personally identifiable information was removed, please refer to the cited papers.}

\begin{itemize}
	\item \textbf{Political books} \cite{polbooks} ($N=105, E=441, D=3$) -- network of Amazon book sales about U.S. politics, published close to the presidential election in 2004. Two books are connected if they were frequently co-purchased by the same customer. Vertex features encode the political affiliation of the author (liberal, conservative or neutral).

	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) -- network of face-to-face contacts amongst students and teachers at a primary school in Lyon, France. Two nodes are connected if the two parties shared a face-to-face interaction over the course of the day. Vertex features include class membership (one of 10 values 1A-5B), gender and whether or not the individual is a teacher or pupil. These data were collected on consecutive days in October 2009. We choose to analyse just the second day.

	\item \textbf{Facebook egonet} \cite{fb-snap} ($N=747, E=30025, D=480$) -- an assortment of Facebook egonets. These are networks of a particular user's friends list and all the connections within that. Vertex features are extracted from each user's profile and are fully anonymized. They include information about education history, languages spoken, gender, home-town, date of birth, name to give a few examples. We focus on the egonet with id 1912.

@@

\begin{figure}[!h]
	\centering
	\begin{subfigure}[t]{0.28\linewidth}
		\centering
		\includegraphics[width=\linewidth]{polbooks-graph.png}
		\caption{Polbooks}
		\label{fig:polbooks-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.28\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-graph.png}
		\caption{School}
		\label{fig:school-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.28\linewidth}
		\centering
		\includegraphics[width=\linewidth]{fb-graph.png}
		\caption{Facebook egonet}
		\label{fig:fb-graph}
	\end{subfigure}
	\begin{subfigure}[t]{0.10\linewidth}
		\centering
		\includegraphics[width=0.8\linewidth]{10-vertical-legend.png}
		\caption{Legend}
		\label{fig:10-legend}
	\end{subfigure}
	\caption{Networks laid out and coloured according by inferred block memberships $\hat{y}$ for a given experiment iteration. Visualisation performed using \textit{graph-tool} \cite{peixoto_graph-tool_2014}}
	\label{fig:graphs-all}
\end{figure}

We require metrics to assess performance. This can be split into two separate components: the microcanonical SBM fit (concerned with the $b$-samples) and the fit of the feature-to-block generator (concerned with the $\theta$-samples). Starting with the SBM, $S(b)$ (equation \ref{eqn:dl-form}) can be interpreted as the description length of the partition imposed by $b$. It is only natural to divide this quantity by the number of entities (nodes and edges) in our graph $N+E$ to allow for rough comparison between graphs. This defines a simple metric to gauge the fit of the SBM: the description length per entity averaged over the $b$-samples (equation \ref{eqn:mean-dl}):
%
\begin{equation}
	\bar{S}_e \coloneqq \frac{1}{(N+E) |\Tcal_b|} \sum_{t\in \Tcal_b} S \left( b^{(t)} \right)
	\label{eqn:mean-dl}
\end{equation}
%
However, to assess the performance of the feature-to-block predictor, we must partition the vertex set $[N]$ into training and test sets: $\Gcal_0$ and $\Gcal_1$ respectively.\footnote{We choose to randomly create the partition on each experiment run such that a constant fraction $f$ of the available vertices go to form our training set $\Gcal_0$ and the remaining vertices are held out to form our test set $\Gcal_1$. For all the experiments $f=0.7$.}
The $b$-chain is run using the whole network but we only use vertices $v \in \Gcal_0$ to train the $\theta$-chain. As $|\Gcal_0| \neq |\Gcal_1|$ sets, we cannot use the un-normalised log target $U$ (equation \ref{eqn:U-form}) for comparison; the total cross-entropy loss is scaled by the size of each set but the prior term stays constant. We therefore must use the average cross-entropy loss over each set (equation \ref{eqn:cross-entropy-loss}):
%
\begin{equation}
	\bar{\Lcal}_\star \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \Lcal_\star \left( \theta^{(t)} \right)
	\qquad \textrm{where} \qquad
	\Lcal_\star \left( \theta^{(t)} \right) \coloneqq \frac{1}{|\Gcal_\star|} \sum_{i \in \Gcal_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)}
	\label{eqn:cross-entropy-loss}
\end{equation}
%
Where $\star \in \{0, 1\}$ has been introduced to toggle between training and test sets. Table \ref{tab:results} summarises the results for each experiment.\footnote{For a comprehensive list of the hyper-parameters used for each experiment please see appendix \ref{appdx:hyperparams}}

We also apply the dimensionality reduction method on the two higher dimensional datasets (the school and FB egonet). For this we leverage equation \ref{eqn:c-star}, to reduce the dimension from $D$ to $D'$ with $k=1$ to yield $c^*$. We then retrain the feature-block predictor using just the retained feature set $\Dcal'$ and report the loss over the training and test sets for the reduced classifier -- denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. These values are also included in table \ref{tab:results}.

\begin{table}[!h]
	\centering
	\caption{Experimental results averaged over $n=10$ iterations (mean $\pm$ standard deviation)}
	\label{tab:results}
	\resizebox{\textwidth}{!}{%
	\begin{tabular}{c|ccc|c|cc|ccc}
		Dataset  & $B$ & $D$ & $D'$ & $\bar{S}_e$ & $\bar{\Lcal}_0$ & $\bar{\Lcal}_1$ & $c^*$ & $\bar{\Lcal}_0'$ & $\bar{\Lcal}_1'$  \\ \hline
		Polbooks & 3 & 3 & -- & $2.250 \pm 0.001$ & $0.584 \pm 0.033$ & $0.557 \pm 0.061$ & -- & -- & -- \\
		School & 10 & 13 & 10 & $1.894 \pm 0.006$ & $0.793 \pm 0.096$ & $0.914 \pm 0.112$ & $1.112 \pm 0.230$ & $0.791 \pm 0.096$ & $0.864 \pm 0.139$ \\
		FB egonet & 10  & 480 & 10 & $1.626 \pm 0.003$ & $1.305 \pm 0.034$ & $1.539 \pm 0.087$ & $0.942 \pm 0.042$ & $1.496 \pm 0.093$ & $1.578 \pm 0.104$
	\end{tabular}
	}
\end{table}

Table \ref{tab:results} already highlights some general trends in the results. Firstly, the variance of the test loss $\bar{\Lcal}_1$ tends to be higher than the training loss $\bar{\Lcal}_0$. This is expected as our test set is smaller than the training set and so more susceptible to variability in its construction. Indeed, much of the variance in the evaluation of $\bar{\Lcal}_0$ and $\bar{\Lcal}_1$ comes from the random partitioning of the graph into training and test sets $\Gcal_0$ and $\Gcal_1$. Secondly, it can be seen that the dimensionality reduction procedure brings the training and test losses closer together. This implies that the features we keep are indeed correlated with the underlying graphical partition.

@@

On figure \ref{fig:null-all} we give the sampled feature weights for the reduced feature-set $\Dcal'$ for a given experiment iteration (or just the original $\Dcal$ for the polbooks dataset). We go on to analyse each of these in turn.

\begin{figure}[!h]
	\centering
	\begin{subfigure}[t]{0.32\linewidth}
		\centering
		\vskip 0pt
		\includegraphics[width=\linewidth]{polbooks-null.png}
		\caption{Polbooks}
		\label{fig:polbooks-null}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.32\linewidth}
		\centering
		\vskip 0pt
		\includegraphics[width=\linewidth]{school-null.png}
		\caption{School}
		\label{fig:school-null}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.32\linewidth}
		\centering
		\vskip 0pt
		\includegraphics[width=\linewidth]{fb-null.png}
		\caption{Facebook egonet}
		\label{fig:fb-null}
	\end{subfigure}
	\caption{Reduced dimension feature-to-block generator weight samples}
	\label{fig:null-all}
\end{figure}

@@

We wish to determine whether political affiliation is a good predictor of the overall network structure. We choose to partition the network into $B=3$ communities as we only have this many distinct values for political affiliation (conservative, liberal or neutral).

From, figure \ref{fig:polbooks-null}, we see that all 3 blocks have a distinct political affiliation as their largest positive component. This is strong evidence that political affiliation is indeed the axis which best predicts the 3-way natural partition of the graph into blocks.

From figure \ref{fig:polbooks-graph}, we note that there are very few edges between blocks 2 and 3 (the detected conservative and liberal blocks as per figure \ref{fig:school-null}). Block 1 effectively acts as a separator between blocks 2 and 3; very few readers venture all the way across the political aisle for their next book.

From table \ref{tab:results} we see that the training and test losses are very similar (the test loss mean is even below the training loss) and both are low in magnitude. This is very strong evidence that the features we have available (the author's political affiliation) are indeed the drivers of the high-level graphical structure at $B=3$.

@@

We must first choose the number of blocks $B$ to define the coarseness of our analysis. A total of 10 school-classes would suggest that $B=10$ is a natural starting point. We visualise the inferred block memberships in figure \ref{fig:school-graph}.

As before, we sample the block-generator parameters $\theta$ and employ the dimensionality reduction technique with standard deviation multiplier $k=1$ to pick out the top $D'=10$ features. We then plot the weights for the surviving features $d \in \Dcal'$ on figure \ref{fig:school-null}. Immediately, we see that only the pupils' class memberships have survived (1A-5B); gender and teacher/student status have been discarded meaning that these are not good predictors of overall macro-structure.

@@

As for the training and test losses of the original and reduced classifiers. We see that the training loss stays roughly constant between the original and reduced classifiers $\bar{\Lcal}_0 \approx \bar{\Lcal}_0'$ but the test loss of the reduced classifier is ever-so-slightly decreased (within margin of error). This improved generalisation implies that the features we have discarded (gender and teacher/student status) do not impact the high-level graphical structure and simply lead to over-fitting in the original classifier. The features we have available do not capture the entire picture but they do quite well (modestly low overall loss).

@@

This dataset was chosen to showcase the power of the dimensionality reduction technique as the feature-space has high dimension ($D=480$). We sample the $b$-chain, specifying $B=10$ total blocks and use this to construct the $\theta$-samples as before.

We again apply the dimensionality reduction technique with $k=1$ to target $D'=10$ features in the reduced set. The remaining features (figure \ref{fig:fb-null}) are those that best explain the high-level community structure. The majority of the surviving features are education related. Nevertheless, for $D'=10$ we only have good explanations for the makeup of some of the detected blocks.

When the feature dimension is very large, it becomes increasingly likely that a particular feature may uniquely identify a small set of nodes. If these nodes are all part of the same community then the classifier will overfit for that particular parameter. The regularisation term imposed by the prior goes some way to alleviating this problem. Nevertheless, we see in figure \ref{fig:fb-null} that the feature \verb*|birthday-5| has a very high weight as it relates to block 1. It might be possible to shift the feature values such that they take values in $\{-1, 1\}$ rather than $\{0, 1\}$ but to accept more than one feature-group per block.

The training and test losses (table \ref{tab:results}) are the highest of the considered networks but nonetheless still low. This suggests that the features we have available only partly explain the high-level structure. Promisingly, after reducing the problem dimension our test loss barely changes, suggesting that the features we do keep are the best from the ones we have available at explaining the overall network structure.

>>>>

We apply the developed methods to a variety of datasets. These are chosen to span a range of node counts $N$, edge counts $E$ and feature space dimension $D$. We consider the following:

\begin{itemize}[leftmargin=*]
	\item \textbf{Political books} \cite{polbooks} ($N=105, E=441, D=3$) -- network of Amazon book sales about U.S. politics, published close to the presidential election in 2004. Two books are connected if they were frequently co-purchased by customers. Vertex features encode the political affiliation of the author (liberal, conservative or neutral).

	\item \textbf{Primary school dynamic contacts} \cite{schools} ($N=238, E=5539, D=13$) -- network of face-to-face contacts amongst students and teachers at a primary school in Lyon, France. Two nodes are connected if the two parties shared a face-to-face interaction over the school-day. Vertex features include class membership (one of 10 values: 1A-5B), gender and teacher/student distinction. No further identifiable information is retained. We choose to analyse just the second day of results.

	\item \textbf{Facebook egonet} \cite{fb-snap} ($N=747, E=30025, D=480$) -- an assortment of Facebook users' friends lists. Vertex features are extracted from each user's profile and are fully anonymised. They include information about education history, languages spoken, gender, home-town, birthday etc. We focus on the egonet with id 1912.

@@

\begin{figure}[!h]
	\centering
	\begin{subfigure}[t]{0.28\linewidth}
		\centering
		\includegraphics[width=\linewidth]{polbooks-graph.png}
		\caption{Polbooks}
		\label{fig:polbooks-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.28\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-graph.png}
		\caption{School}
		\label{fig:school-graph}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.28\linewidth}
		\centering
		\includegraphics[width=\linewidth]{fb-graph.png}
		\caption{Facebook egonet}
		\label{fig:fb-graph}
	\end{subfigure}
	\begin{subfigure}[t]{0.10\linewidth}
		\centering
		\includegraphics[width=0.8\linewidth]{10-vertical-legend.png}
		\caption{Legend}
		\label{fig:10-legend}
	\end{subfigure}
	\caption{Networks laid out and coloured according by inferred block memberships $\hat{y}$ for a given experiment iteration. Visualisation performed using \textit{graph-tool} \cite{peixoto_graph-tool_2014}}
	\label{fig:graphs-all}
\end{figure}
\begin{figure}[!h]
	\centering
	\begin{subfigure}[t]{0.32\linewidth}
		\centering
		\vskip 0pt
		\includegraphics[width=\linewidth]{polbooks-null.png}
		\caption{Polbooks}
		\label{fig:polbooks-null}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.32\linewidth}
		\centering
		\vskip 0pt
		\includegraphics[width=\linewidth]{school-null.png}
		\caption{School}
		\label{fig:school-null}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.32\linewidth}
		\centering
		\vskip 0pt
		\includegraphics[width=\linewidth]{fb-null.png}
		\caption{Facebook egonet}
		\label{fig:fb-null}
	\end{subfigure}
	\caption{Reduced dimension feature-to-block generator weight samples}
	\label{fig:null-all}
\end{figure}
\begin{table}[!h]
	\centering
	\caption{Experimental results averaged over $n=10$ iterations (mean $\pm$ standard deviation)}
	\label{tab:results}
	\resizebox{\textwidth}{!}{%
		\begin{tabular}{c|ccc|c|cc|ccc}
			Dataset  & $B$ & $D$ & $D'$ & $\bar{S}_e$ & $\bar{\Lcal}_0$ & $\bar{\Lcal}_1$ & $c^*$ & $\bar{\Lcal}_0'$ & $\bar{\Lcal}_1'$  \\ \hline
			Polbooks & 3 & 3 & -- & $2.250 \pm 0.001$ & $0.584 \pm 0.033$ & $0.557 \pm 0.061$ & -- & -- & -- \\
			School & 10 & 13 & 10 & $1.894 \pm 0.006$ & $0.793 \pm 0.096$ & $0.914 \pm 0.112$ & $1.112 \pm 0.230$ & $0.791 \pm 0.096$ & $0.864 \pm 0.139$ \\
			FB egonet & 10  & 480 & 10 & $1.626 \pm 0.003$ & $1.305 \pm 0.034$ & $1.539 \pm 0.087$ & $0.942 \pm 0.042$ & $1.496 \pm 0.093$ & $1.578 \pm 0.104$
		\end{tabular}
	}
\end{table}
%
We require metrics to assess performance. This can be split into two separate components: the microcanonical SBM fit (concerned with the $b$-samples) and the fit of the feature-to-block generator (concerned with the $\theta$-samples). Starting with the SBM, $S(b)$ (equation \ref{eqn:dl-form}) can be interpreted as the description length of the partition imposed by $b$. It is only natural to divide this quantity by the number of entities (nodes and edges) in our graph $N+E$ to allow for rough comparison between graphs. This defines a simple metric to gauge the fit of the SBM: the description length per entity averaged over the $b$-samples (equation \ref{eqn:mean-dl}):
%
\begin{equation}
	\bar{S}_e \coloneqq \frac{1}{(N+E) |\Tcal_b|} \sum_{t\in \Tcal_b} S \left( b^{(t)} \right)
	\label{eqn:mean-dl}
\end{equation}
%
However, to assess the performance of the feature-to-block predictor, we must partition the vertex set $[N]$ into training and test sets. We choose to randomly partition the vertices on each experiment run such that a constant fraction $f$ of the available vertices go to form our training set $\Gcal_0$ and the remainder are held out to form our test set $\Gcal_1$.
The $b$-chain is run using the whole network but we only use vertices $v \in \Gcal_0$ to train the $\theta$-chain. As $|\Gcal_0| \neq |\Gcal_1|$ in general, we cannot use the un-normalised log target $U$ (equation \ref{eqn:U-form}) for comparison as the total cross-entropy loss is scaled by the size of each set but the prior term stays constant. We therefore must use the average cross-entropy loss over each set (equation \ref{eqn:cross-entropy-loss}):
%
\begin{equation}
	\bar{\Lcal}_\star \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \Lcal_\star \left( \theta^{(t)} \right)
	\qquad \textrm{where} \qquad
	\Lcal_\star \left( \theta^{(t)} \right) \coloneqq \frac{1}{|\Gcal_\star|} \sum_{i \in \Gcal_\star}\sum_{j \in [B]} \hat{y}_{ij} \log \frac{1}{\phi_j \left(x_i; \theta^{(t)} \right)}
	\label{eqn:cross-entropy-loss}
\end{equation}
%
Where $\star \in \{0, 1\}$ has been introduced to toggle between training and test sets. Table \ref{tab:results} summarises the results for each experiment.\footnote{For a comprehensive list of the hyper-parameters used for each experiment please see appendix \ref{appdx:hyperparams}} We also apply the dimensionality reduction method on the two higher dimensional datasets (the school and FB egonet). For this we leverage equation \ref{eqn:c-star}, to reduce the dimension from $D$ to $D'$ with $k=1$ to yield the maximal cutoff $c^*$. We then retrain the feature-block predictor using just the retained feature set $\Dcal'$ and report the loss over the training and test sets for the reduced classifier -- denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. These values are also included in table \ref{tab:results}.

Table \ref{tab:results} already highlights some general trends in the results. Firstly, the variance of the test loss $\bar{\Lcal}_1$ tends to be higher than the training loss $\bar{\Lcal}_0$. This is expected as our test set is smaller than the training set and so more susceptible to variability in its construction. Indeed, most of the variance in the evaluation of $\bar{\Lcal}_0$ and $\bar{\Lcal}_1$ comes from the random partitioning of the graph into training and test sets. Secondly, it can be seen that the dimensionality reduction procedure brings the training and test losses closer together. This implies that the features we keep are indeed correlated with the underlying graphical partition and that the approach generalises correctly.

@@

@@

We wish to determine whether the author's political affiliation is a good predictor of the overall network structure. We choose to partition the network into $B=3$ communities as we only have this many distinct values for political affiliation (conservative, liberal or neutral). From, figure \ref{fig:polbooks-null}, we see that all 3 blocks have a distinct political affiliation as their largest positive component. This is strong evidence that political affiliation is indeed the axis which best predicts the 3-way natural partition of the graph into blocks. Furthermore, table \ref{tab:results} we see that the training and test losses are very similar (the test loss mean is even below the training loss) and both are low in magnitude. This provides further evidence to the claim that political affiliation is the best explanatory variable for the overall network structure.

@@

We choose $B=10$ in line with the total number of school-classes. As before, we sample the block-generator parameters $\theta$ and employ the dimensionality reduction technique with standard deviation multiplier $k=1$ to pick out the top $D'=10$ features. We then plot the weights for the surviving features $d \in \Dcal'$ on figure \ref{fig:school-null}. Immediately, we see that only the pupils' class memberships have survived (1A-5B); gender and teacher/student status have been discarded meaning that these are not good predictors of overall macro-structure.

@@

@@

We choose $B=10$ and $D'=10$ for this experiment. The remaining features (figure \ref{fig:fb-null}) are those that best explain the high-level community structure. The majority of the surviving features are education related. Nevertheless, for $D'=10$ we only have good explanations for the makeup of some of the detected blocks; several blocks in figure \ref{fig:fb-null} do not have high-magnitude components for $D'=10$.

When the feature dimension is very large, it becomes increasingly likely that a particular feature may uniquely identify a small set of nodes. If these nodes are all part of the same community then the classifier will overfit for that particular parameter. The regularisation term imposed by the prior goes some way to alleviating this problem. Nevertheless, we see in figure \ref{fig:fb-null} that the feature \verb*|birthday-5| has a very high weight as it relates to block 1 -- but it would be preposterous to conclude that birthdays determine graphical structure. The analyst must remain vigilant of such problems.

### 2021-05-23 nips/ffbm.tex
Finish draft

In this section we propose a novel generative model for labelled networks. As before, we let $N$ denote the number of nodes. We define the vector $x_i \in \Xcal^D$ as the feature vector for the $i$'th vertex. Each vertex has $D$ total features and we assume all entries take values from the same set $\Xcal$. For the datasets we analyse, we deal with binary feature flags so $\Xcal = \{0, 1\}$. The feature vectors $\{x_i\}_{i=1}^{N}$ may be compactly subsumed into the feature matrix $X \in \Xcal^{N \times D}$.

The proposed generative model (which we call the feature-first block model - FFBM) is given in figure \ref{fig:ffbm}. First, we start with the feature matrix X and probabilistically generate a vector of block memberships $b$. The parameters of this step are encapsulated by $\theta$. Each feature vector $x_i$ is treated independently and used to generate the corresponding block membership $b_i$. Each entry $b_i \in [B]$ where $B$ is the number of blocks and a constant in our model. We choose a single softmax layer to model $p(b_i | x_i, \theta)$. More complex models are possible but then deriving meaning from the inferred parameter distributions is more difficult. Summarising, we write $p(b | X, \theta)$ as follows:
%
\begin{equation}
	p(b| X, \theta) = \prod_{i=1}^{N} p(b_i | x_i, \theta) = \prod_{i=1}^{N} \phi_{b_i} (x_i; \theta)
	= \prod_{i=1}^{N} \frac{\exp\left(w_{b_i}^T \tilde{x}_i\right)}{\sum_{k=1}^{B} \exp \left( w_k^T \tilde{x}_i\right)}
\end{equation}
%
Where $\tilde{x} \coloneqq \left[x_1, x_2, \dots x_D, 1\right]^T$ is an augmented version of $x$ that allows for a bias term. The parameter vector $\theta$ for this stage contains of the all the weight vectors $\theta = \{w_k\}_{k=1}^{B}$. Each $w_k$ has dimension $D+1$. We could instead write the parameters $\theta$ as a $B \times (D+1)$ matrix of weights $W$; this form has computational benefits as then $z_i \coloneqq W \tilde{x}_i$ is the input to the softmax activation function.

@@

		% arrows
		\draw[->] (X.east) -- node[above] {$\theta$} (b.west);
		\draw[->] (b.east) -- node[above] {$\psi$}(A.west);
	\end{tikzpicture}
	\caption{Feature-first block model}
	\label{fig:ffbm}
\end{figure}

Once the block memberships $b$ have been generated, we then draw the graph $A$ from the microcanonical DC-SBM (equation \ref{eqn:A-generation}) with additional parameters encapsulated by $\psi = \{e, k\}$. In a slight abuse of notation we denote the inter-block edge count matrix with $e = \psi_e$ and the degree sequence $k = \psi_k$ to make explicit that these parameters are contained in $\psi$.
%
\begin{equation}
	A \sim \textrm{DC-SBM}_{\textrm{MC}} (b, \psi_e, \psi_k)
	\label{eqn:A-generation}
\end{equation}

@@

Before performing any inference, we must specify priors on $\theta$ and $\psi$. For $\theta$ it seems sensible to choose a Gaussian prior, with zero mean and variance matrix $\sigma^2_\theta I$ such that each element of $\theta$ is independent and distributed like $\sim \Gaussian(0, \sigma_\theta^2)$. In vector form, the prior for $\theta$ is therefore:
%
\begin{equation}
	p(\theta) = \Gaussian \left( \theta ; 0, \sigma_\theta^2 I \right)
	\label{eqn:theta-prior}
\end{equation}
%
We will see that this form of prior is equivalent to a regularisation term in neural network training that penalises extreme weight magnitudes. For $\sigma_\theta^2 \rightarrow \infty$ this becomes an uninformative uniform prior.

In our model, the block memberships vector $b$ is an intermediate latent variable and so we are not free to choose a prior for it. Nevertheless, as far as inference on the right-hand-side of figure \ref{fig:ffbm}, we regard $p(b | X)$ as a pseudo-prior on $b$. We can show (appendix \ref{appdx:b|x}) that our choice of prior for $p(\theta)$ in equation \ref{eqn:theta-prior} leads to a uniform $p(b | X)$ in equation \ref{eqn:b-pseudo-prior}.
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}
	\label{eqn:b-pseudo-prior}
\end{equation}
%
This is a uniform distribution that only depends on the number of blocks $B$. This is an enormously important simplification as evaluating $p(b | X)$ does not require knowing the exact value $X$ takes and can be performed without an expensive Monte-Carlo integration over the $\theta$-domain. \citet{Peixoto-Bayesian-Microcanonical} proposes careful choices for the additional microcanonical SBM parameters $\psi$ which we adopt. Peixoto's idea is to write the joint prior on $(b, e, k)$ as a product of conditionals $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. For our purposes we must insert a conditioning on $X$, to form our pseudo-prior for $b$ and $\psi$, to give equation \ref{eqn:joint-pseudo-prior}.
%
\begin{equation}
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b)
	\label{eqn:joint-pseudo-prior}
\end{equation}
%
Where the simplification is made apparent by noting $(\psi \indep X) | b$. We then borrow the priors proposed by \citet{Peixoto-Bayesian-Microcanonical} for $p(\psi | b)$ to complete our model. Please refer to appendix \ref{appdx:prior} for the exact form of $p(\psi | b)$. All that concerns us is that we have a form which is computable.

>>>>

In this section we propose a novel generative model for labelled networks. We call this the feature-first block model (FFBM) and outline its structure in \ref{fig:ffbm} As before, we let $N$ denote the number of nodes and $B$ the number of blocks in our graph. We define the vector $x_i \in \Xcal^D$ as the feature vector for the $i$'th vertex. $D$ is the number of features. For the datasets we analyse, we deal with binary feature flags so $\Xcal = \{0, 1\}$. The feature vectors $\{x_i\}_{i=1}^{N}$ may be compactly subsumed into the feature matrix $X \in \Xcal^{N \times D}$.

For the FFBM, we start with the feature matrix X and probabilistically generate a vector of block memberships $b \in [B]^N$. The parameters of this step are encapsulated by $\theta$. Each feature vector $x_i$ is treated independently and used to generate the corresponding block membership $b_i \in [B]$. We choose a single softmax layer to model $p(b_i | x_i, \theta)$. More complex models are possible but then deriving meaning from the inferred parameter distributions is more difficult. Summarising, we write $p(b | X, \theta)$ as follows:
%
\begin{equation}
	p(b| X, \theta) = \prod_{i=1}^{N} p(b_i | x_i, \theta) = \prod_{i=1}^{N} \phi_{b_i} (x_i; \theta)
	= \prod_{i=1}^{N} \frac{\exp\left(w_{b_i}^T \tilde{x}_i\right)}{\sum_{k=1}^{B} \exp \left( w_k^T \tilde{x}_i\right)}
\end{equation}
%
Where $\tilde{x} \coloneqq \left[x_1, x_2, \dots x_D, 1\right]^T$ is an augmented version of $x$ that allows for a bias term. The parameter vector $\theta$ for this stage contains all the weight vectors $\theta = \{w_k\}_{k=1}^{B}$. Each $w_k$ has dimension $D+1$. We could instead write the parameters $\theta$ as a $B \times (D+1)$ matrix of weights $W$; this form has computational benefits as then $z_i \coloneqq W \tilde{x}_i$, which is the input to the softmax activation function.

@@

		% arrows
		\draw[->] (X.east) -- node[above] {$\theta$} (b.west);
		\draw[->] (b.east) -- node[above] {$\psi$}(A.west);
	\end{tikzpicture}
	\caption{The feature-first block model (FFBM)}
	\label{fig:ffbm}
\end{figure}

Once the block memberships $b$ have been generated, we then draw the graph $A$ from the microcanonical DC-SBM (equation \ref{eqn:A-generation}) with additional parameters encapsulated by $\psi = \{\psi_e, \psi_k\}$.
%
\begin{equation}
	A \sim \textrm{DC-SBM}_{\textrm{MC}} (b, \psi_e, \psi_k)
	\label{eqn:A-generation}
\end{equation}

@@

Before performing any inference, we must specify priors on $\theta$ and $\psi$. For $\theta$ it seems sensible to choose a Gaussian prior, with zero mean and variance matrix $\sigma^2_\theta I$ such that each element of $\theta$ is independent and distributed like $\sim \Gaussian(0, \sigma_\theta^2)$. In vector form, the prior for $\theta$ is therefore:
%
\begin{equation}
	p(\theta) = \Gaussian \left( \theta ; 0, \sigma_\theta^2 I \right)
	\label{eqn:theta-prior}
\end{equation}
%
In our model, the block memberships vector $b$ is an intermediate latent variable and so we are not free to choose a prior for it. Nevertheless, as far as inference on the right-hand-side of figure \ref{fig:ffbm}, we regard $p(b | X)$ as a pseudo-prior on $b$. We can show (appendix \ref{appdx:b|x}) that our choice of prior for $p(\theta)$ in equation \ref{eqn:theta-prior} leads to a uniform $p(b | X)$ in equation \ref{eqn:b-pseudo-prior}.
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}
	\label{eqn:b-pseudo-prior}
\end{equation}
%
This is an enormously important simplification as evaluating $p(b | X)$ does not require an expensive Monte-Carlo integration over the $\theta$-domain nor does it require the exact value of $X$. \citet{Peixoto-Bayesian-Microcanonical} proposes careful choices for the additional microcanonical SBM parameters $\psi$ which we adopt. Peixoto's idea is to write the joint prior on $(b, e, k)$ as a product of conditionals $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. For our purposes we must insert a conditioning on $X$, to form our pseudo-prior for $b$ and $\psi$, to give equation \ref{eqn:joint-pseudo-prior}.
%
\begin{equation}
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b)
	\label{eqn:joint-pseudo-prior}
\end{equation}
%
Where we leverage the fact $(\psi \indep X) | b$. We then borrow the priors proposed by \citet{Peixoto-Bayesian-Microcanonical} for $p(\psi | b)$ to complete our model. Please refer to appendix \ref{appdx:prior} for the exact form of $p(\psi | b)$. All that concerns the main argument is we have a computable form.
